Staged Abstract Interpreters

### 2019-09-08 paper/30_stageconc.tex
polish

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to stage the concrete
interpreter by changing the monad type and refactoring primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala, and then replay the same steps from the unstaged
counterpart. Finally, we briefly describe the code generation.

@@

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} which
use syntactic quotations, LMS distinguishes binding-time based on types. 
LMS provides a type constructor @Rep[T]@ where @T@ can be
an arbitrary type, indicating a value of type @T@ will be known at the future stage.
Operations acting on @Rep[T]@ expressions will be residualized as generated
code.
A classic example for introducing multi-stage programming is the power function
that computes $b^x$. Usually, it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

>>>>

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to stage the concrete
interpreter by changing the monad type and refactoring the primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala, and then replay the same steps from the unstaged
counterpart. Finally, we briefly describe the code generation.

@@

Lightweight Modular Staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} which
uses syntactic quotations, LMS distinguishes binding-times based on types. 
LMS provides a type constructor @Rep[T]@, where @T@ can be an arbitrary type,
indicating a value of type @T@ will be known at the future stage.
Operations acting on @Rep[T]@ expressions will be residualized as generated
code.
A classic example for introducing multi-stage programming is the power function
that computes $b^x$. Usually, it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

### 2019-09-08 paper/60_cases.tex
space
revision
polish

In Section~\ref{sai}, we used a toy language to show that staging an abstract
interpreter is feasible and can be systematically derived from staged concrete
interpreters and unstaged abstract interpreters.
In this section, we conduct several case studies to further show that this
methodology is practically useful and widely applicable to diverse analyses.

@@

In this section, we show AC (closure generation) can be understood and
implemented as an instance of staging abstract interpreters. We first revisit
the original implementation of abstract compilation for 0-CFA, and then
reproduce their result by simply adding stage annotations.
The program generated by our approach provides approximately the same
improvement with regards to speed, but without changing a single line of the
analyzer program using type-based staging annotations. However, AC requires
more engineering effort, i.e., rewriting the whole analyzer into the
closure-generation form. Moreover, as shown in Section~\ref{staged_ds}, our
approach is able to not only remove the interpretive overhead, but also enable
several optimizations such as specialized data structures.

@@

  \end{subfigure}
  \caption{Comparison of AC (left) and SAI (right). Only core code are shown.}
  \label{compare_ac_sai}
\end{figure*}

@@

\begin{figure}[h!]
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out
    val res: Rep[(Set[(Value,Store)], Cache)] = ...  // omit code for checking caches
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2));     _ <- put_store(σ)  // gc performs abstract garbage collection
  } yield v
\end{lstlisting}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\end{figure}

>>>>

In Section~\ref{sai}, we used a toy language to show that staging an abstract
interpreter is feasible and the artifact can be derived by combining the staged
concrete interpreter and unstaged abstract interpreter.
In this section, we conduct several case studies to further show that this
methodology is practically useful and widely applicable to diverse analyses.

@@

In this section, we show AC can be understood and
implemented as an instance of staging abstract interpreters. We first revisit
the original implementation of abstract compilation for 0-CFA, and then
reproduce their result by simply adding stage annotations.
The program generated by our approach provides approximately the same
improvement with regards to speed, but without changing a single line of the
analyzer program using type-based staging annotations. However, AC requires
more engineering effort, i.e., rewriting the whole analyzer into the
closure-generation form. Moreover, as shown in Section~\ref{staged_ds}, our
approach is able to not only remove the interpretive overhead, but also enable
several optimizations such as specialized data structures.

@@

  \end{subfigure}
  \caption{Comparison of AC (left) and SAI (right). Only core code are shown.}
  \vspace{-1.5em}
  \label{compare_ac_sai}
\end{figure*}

@@

\begin{figure}[h!]
\vspace{-1em}
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out
    val res: Rep[(Set[(Value,Store)], Cache)] = ...  // omit code for checking caches
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2));     _ <- put_store(σ)  // gc performs abstract garbage collection
  } yield v
\end{lstlisting}
\vspace{-0.5em}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\vspace{-2em}
\end{figure}

### 2019-09-08 paper/80_related.tex
misc
polish

\paragraph{Construction of Abstract Interpreters} Abstract interpretation is a
semantics-based approach to build sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77, Cousot98-5}.
As for building semantic artifacts, the Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach shows that
abstract interpreters can be derived systematically from concrete semantic
artifacts based on operational semantics.
The AAM approach is closely related to control-flow analysis for higher-order languages
\cite{Midtgaard:2012:CAF:2187671.2187672, Shivers:1991:SSC:115865.115884}.
Using monads to construct abstract interpreters was explored by
\citet{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17,
Darais:2015:GTM:2814270.2814308}.
The unstaged abstract interpreter in this paper follows the abstracting
definitional interpreters (ADI) approach from
\citet{DBLP:journals/pacmpl/DaraisLNH17}.  Similar to ADI, reconstructing
big-step abstract interpreters with delimited continuations was shown by
\citet{Wei:2018:RAA:3243631.3236800}; \citet{Keidel:2018:CSP:3243631.3236767}
presented a modular, arrow-based abstract interpreter that allows soundness
proof to be constructed compositionally.
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis, where we borrow the notation of modular analysis.
However, the staged abstract interpreter does not generate complete summaries for
modules; instead, it is able to specialize the analysis modularly. Similarly, AC
is considered as an example of this kind of modular analyses \cite{DBLP:conf/cc/CousotC02}.

@@

\paragraph{Two-level Semantics} The idea of reinterpreting the semantics as
abstract interpretation can be traced to Nielson's two-level semantics
\cite{NIELSON1989117}; using two-level semantics for code generation
was also explored by \citet{NIELSON198859}.
\citet{Sergey:2013:MAI:2491956.2491979}'s monadic abstract
interpreters is closely related to the two-level semantics: the
use of a generic interface with monads and then reinterpreting it with
different semantics is already two-level. Instead of focusing on
semantics, this paper shows how a staged analyzer can be built and
used to increase efficiency of static analysis. We augment the monadic
abstract interpreter from another dimension, using code generation to
produce efficient low-level code. Hence the presented work can be
considered as a two-dimensional application of the two-level semantics.

\paragraph{Partial Evaluation and Multi-stage Programming}
Partial evaluation is an automatic technique for program
specialization and was studied comprehensively by
\citet{10.1007/3-540-61580-6_11, DBLP:books/daglib/0072559}.
In this paper, we use mutli-stage programming to specialize programs.
The Lightweight Modular Staging (LMS) framework
\cite{DBLP:conf/gpce/RompfO10} we used in the paper relies on
type-level stage annotations.  Other notable MSP systems using syntactic
quotations exist in the ML family, e.g., MetaML \cite{DBLP:conf/pepm/TahaS97}
and MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}. 
Compared with other approaches using syntactic quotations, LMS provides
guarantee for preserving evaluation orders within a stage.
\citet{Ofenbeck:2017:SGP:3136040.3136060} discussed staging for generic
programming as well as stage polymorphism in LMS.
%The idea of staging an abstract interpreter we presented here
%is also applicable to other MSP systems.
Multi-stage programming has been widely used to improve the
performance in many other domains, such as optimizing compilers or
domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14,
  DBLP:conf/snapl/RompfBLSJAOSKDK15,
  DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
  DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
  DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17,
  Ofenbeck:2017:SGP:3136040.3136060}, data processing
\cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
  DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.

>>>>

\paragraph{Construction of Abstract Interpreters} Abstract interpretation is a
semantics-based approach to build sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77, Cousot98-5}.
As for building semantic artifacts, the Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach shows that
abstract interpreters can be derived systematically from concrete semantic
artifacts based on operational semantics.
The AAM approach is closely related to control-flow analysis for higher-order languages
\cite{Midtgaard:2012:CAF:2187671.2187672, Shivers:1991:SSC:115865.115884}.
Using monads to construct abstract interpreters was explored by
\citet{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17,
Darais:2015:GTM:2814270.2814308}.
The unstaged abstract interpreter in this paper follows the abstracting
definitional interpreters (ADI) approach \cite{DBLP:journals/pacmpl/DaraisLNH17}.  
Similar to ADI, reconstructing
big-step abstract interpreters with delimited continuations was shown by
\citet{Wei:2018:RAA:3243631.3236800}; \citet{Keidel:2018:CSP:3243631.3236767}
presented a modular, arrow-based abstract interpreter that allows soundness
proof to be constructed compositionally.
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis, from where we borrow the notation of modular analysis.
However, the staged abstract interpreter does not generate complete summaries for
modules; instead, it specializes the analysis modularly. Similarly, AC
is considered as an example of this kind of modular analyses \cite{DBLP:conf/cc/CousotC02}.

@@

\paragraph{Two-level Semantics} The idea of reinterpreting the semantics as
abstract interpretation can be traced to Nielson's two-level semantics
\cite{NIELSON1989117}; using two-level semantics for code generation
was also explored by \citet{NIELSON198859}.
\citet{Sergey:2013:MAI:2491956.2491979}'s monadic abstract
interpreters is closely related to the two-level semantics: the
use of a generic interface with monads and then reinterpreting it with
different semantics is already two-level. Instead of focusing on
semantics, this paper shows how a staged analyzer can be built and
used to increase efficiency of static analyses. We augment the monadic
abstract interpreter with binding-time polymorphism, using code generation to
produce efficient low-level code. The presented work can be considered as
a two-dimensional application of the two-level semantics.

\paragraph{Partial Evaluation and Multi-stage Programming}
Partial evaluation is an automatic technique for program
specialization and was studied comprehensively by
\citet{10.1007/3-540-61580-6_11, DBLP:books/daglib/0072559}.
In this paper, we use mutli-stage programming to specialize programs.
The Lightweight Modular Staging (LMS) framework
\cite{DBLP:conf/gpce/RompfO10} we used in the paper relies on
type-level stage annotations.  Other notable MSP systems using syntactic
quotations exist in the ML family, e.g., MetaML \cite{DBLP:conf/pepm/TahaS97}
and MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}. 
Compared with other approaches using syntactic quotations, LMS provides
guarantee for preserving evaluation orders within a stage.
\citet{Ofenbeck:2017:SGP:3136040.3136060} and
\citet{Amin:2017:CTI:3177123.3158140} discussed staging for generic programming
as well as stage polymorphism in LMS.
%The idea of staging an abstract interpreter we presented here
%is also applicable to other MSP systems.
Multi-stage programming has been widely used to improve the
performance in many other domains, such as optimizing compilers or
domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14,
  DBLP:conf/snapl/RompfBLSJAOSKDK15,
  DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
  DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
  DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17,
  Ofenbeck:2017:SGP:3136040.3136060}, data processing
\cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
  DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.
